---
title: "Intermediate HW"
author: "Lexie DelViscio"
date: "1/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Loop to be evaluated in #1-3, and modified in #4

```{r}
sp_ids <- unique(iris$Species)

output <- matrix(0, nrow=length(sp_ids), ncol=ncol(iris)-1)
rownames(output) <- sp_ids
colnames(output) <- names(iris[ , -ncol(iris)])

for(i in seq_along(sp_ids)) {
    iris_sp <- subset(iris, subset=Species == sp_ids[i], select=-Species)
    for(j in 1:(ncol(iris_sp))) {
        x <- 0
        y <- 0
        if (nrow(iris_sp) > 0) {
            for(k in 1:nrow(iris_sp)) {
                x <- x + iris_sp[k, j]
                y <- y + 1
            }
            output[i, j] <- x / y 
        }
    }
}
output
```

1) Describe the values stored in the object output. In other words what did the loops create?

The values stored in the object output are the mean values of the length and width of the petals for each Iris species (Setosa, Versicolor, and Virginica). The loops created a matrix with 3 rows corresponding with the species name and 4 columns corresponding with the original iris columns, without the species column. In addition, the data contained in these rows is once again the averages for each species. 

2) Describe using pseudo-code how output was calculated

loop from 1 through the length of species identities 
    subset of iris data without the species column
    loop from 1 through the number of columns in the iris_sp data
    if the number of rows in the iris_sp data is greater than 0
      loop from 1 through the number of rows in the iris_sp data
          make x a value equal to 0 + the value in the iris_sp matrix 
          then make y equal to y + 1 (the total number of rows in the matrix + 1)
          the values of output are then set equal to x/y which are the averages 
output is then printed and it's contents were explained above.

3) The variables in the loop were named so as to be vague. How can the objects 
output, x, and y be renamed such that it is clearer what is occurring in the loop.

x is the cumulative values of the data iris_sp for each unique column and row, it
would make sense to rename this variable as cumulative

y can be described as the total frequency of the unique values,
so in relation to statistics it would make sense to rename y as n. Since this is 
a common variable used when discussing sample size of data. 
          
output can be renamed as average_by_species as this is what the matrix itself is
showing. The species names are each of the rows, and each of the columns is a 
specific measurement. Contained in the matrix is the average values calculated 
from the complete data set. So, this should be renamed as average_by_species so 
that it reflects what is being shown. 

4) It is possible to accomplish the same task using fewer lines of code? 
Please suggest one other way to calculate output that decreases the number of loops by 1.

```{r}
sp_ids <- unique(iris$Species)

average_by_species <- matrix(0, nrow=length(sp_ids), ncol=ncol(iris)-1)
rownames(average_by_species) <- sp_ids
colnames(average_by_species) <- names(iris[ , -ncol(iris)])
cumulative <- NULL

for(names in seq_along(sp_ids)) {
    iris_sp <- subset(iris, subset=Species == sp_ids[names], select=-Species)
    for(n in names(iris_sp)) {
      if(class(iris_sp[,n]) == 'integer' | class(iris_sp[,n]) == 'numeric'){
        cumulative <- mean(iris_sp[,n])
          }
          average_by_species[names, n] <- cumulative
        }
    }
average_by_species



```
 

Sum of a sequence
5) You have a vector x with the numbers 1:10. Write a for loop that will produce 
a vector y that contains the sum of x up to that # index of x. So for example the 
elements of x are 1, 2, 3, and so on and the elements of y would be 1, 3, 6, and so on.

```{r}
x <- 1:10
y <- NULL
for(i in x) {
    y <- x * (x+1) / 2
}
y
```


6) Modify your for loop so that if the sum is greater than 10 the value of y is set to NA

```{r}
x <- 1:10
y <- NULL
for(i in x) {
    sum <- x * (x+1) / 2
    if(sum[i] <= 10){
      y <- sum
    }
    else{
      y[i] <- NA
    }
}
y
```

7) Place your for loop into a function that accepts as its argument any vector of 
arbitrary length and it will return y.

```{r}
sum_of_sequence <- function(x) {
  y <- NULL
  x <- 1:x
  for(i in x) {
    sum <- x * (x+1) / 2
    if(sum[i] <= 10){
      y <- sum
    }
    else{
      y[i] <- NA
    }
  }
  return(y)
}

sum_of_sequence(18)
```


