---
title: "Multivariate Assignment"
author: "Lexie DelViscio and Francesca Dellacqua"
date: "2/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Multivariate Assignment 
```{r}
library(vegan) 
data(dune)
data(dune.env)
head(dune)
head(dune.env)
names(dune) # species 
names(dune.env) # variable
```
# 1) Describe how you interpret the graphic. What is the goal of creating such a plot? Does this analysis suggest any interesting findings with respect to the dune vegetation?

```{r}
par(mfrow=(c(1,2)))
dune_mds = metaMDS(dune)

plot(dune_mds, type='n')
text(dune_mds, 'sp', cex=.5)
# generate vector of colors 
color_vect = rev(terrain.colors(6))[-1]
points(dune_mds, 'sites', pch=19, 
       col=color_vect[dune.env$Moisture])
legend('topright', paste("Moisture =", 1:5, sep=''), 
       col=color_vect, pch=19)

dune_mds
#str(dune_mds)
```
the first axis is broader than the second
and, the first variable matches up with the first axis pretty well.

This graphic shows observations of different species of dune vegetation in correlation with moisture levels (n=1-5) based on a non-metric multidimensional scaling model. The goal of creating this plot was to model out these observations as to identify how the moisture levels are correlated with the separate dune vegetation species. This analysis does suggest that certain species are much more correlated with higher moisture levels which is important to examine and understand before conducting other model analyses with added explanatory variables for vegetation distribution and variability in growth.

Species Eleopalu and Ranuflam are usually found close to one another in an area with a lot of higher level moisture points,
whereas Achmill and Triforat are concentrated around lots of low moisture points

it seems like species on the left tend to line up more with lower levels of moisture, 
species towards the right of the plot line up more with higher levels pf moisture.
showing there may be a strong correlation between moisture and where each specific species is found. However, some species 
appear without ant moisture level near them indicating not all variance is explained by moisture. 

note. it computes between every sample a distance matrix, ignores where both are missing since it is a non euclidean method 

# 2)Carry out a direct ordination using CCA in order to test any potential hypotheses that you developed after examining the MDS plot. 

Canonical Correspondence Analysis (CCA)
Let's carry out a Canonical Correspondence Analysis (CCA) as well. CCA is appropriate 
for modeling unimodal or hump-shaped responses to explanatory variables (rather than linear as with RDA).

```{r}
cca_dune <- cca(dune ~ ., data=dune.env)
RsquareAdj(cca_dune, 100)
anova(cca_dune, permutations = 999)
anova(cca_dune, by='margin', permutations = 999)

plot(cca_dune, type='n', scaling=1)
orditorp(cca_dune, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune, display='cn', col='red')
```
just moisture
```{r}
cca_dune_moisture <- cca(dune ~ dune.env$Moisture)
RsquareAdj(cca_dune_moisture, 100)
anova(cca_dune_moisture, permutations = 999)
anova(cca_dune_moisture, by='margin', permutations = 999)

plot(cca_dune_moisture, type='n', scaling=1)
orditorp(cca_dune_moisture, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune_moisture, display='cn', col='red')
dune.env$Moisture
levels(dune.env$Moisture)
```
Just A1
```{r}
cca_dune_A1 <- cca(dune ~ dune.env$A1)
RsquareAdj(cca_dune_A1, 100)
anova(cca_dune_A1, permutations = 999)
anova(cca_dune_A1, by='margin', permutations = 999)

plot(cca_dune_A1, type='n', scaling=1)
orditorp(cca_dune_A1, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune_A1, display='bp', col='red')
```
Just manure
```{r}
cca_dune_Manure <- cca(dune ~ dune.env$Manure)
RsquareAdj(cca_dune_Manure, 100)
anova(cca_dune_Manure, permutations = 999)
anova(cca_dune_Manure, by='margin', permutations = 999)

plot(cca_dune_Manure, type='n', scaling=1)
orditorp(cca_dune_Manure, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune_Manure, display='cn', col='red')
```
Just Management
```{r}
cca_dune_Management <- cca(dune ~ dune.env$Management)
RsquareAdj(cca_dune_Management, 100)
anova(cca_dune_Management, permutations = 999)
anova(cca_dune_Management, by='margin', permutations = 999)

plot(cca_dune_Management, type='n', scaling=1)
orditorp(cca_dune_Management, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune_Management, display='cn', col='red')
```
Just Use
```{r}
cca_dune_Use <- cca(dune ~ dune.env$Use)
RsquareAdj(cca_dune_Use, 100)
anova(cca_dune_Use, permutations = 999)
anova(cca_dune_Use, by='margin', permutations = 999)

plot(cca_dune_Use, type='n', scaling=1)
orditorp(cca_dune_Use, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune_Use, display='cn', col='red')
```
vector moisture - the following two variables were changed to vectors in order to further analyze their individual/overall effects
```{r}
vector_moisture <- as.numeric(as.character(dune.env$Moisture))
vector_moisture
dune.env$Moisture
vector_moisture_cca <- cca(dune~vector_moisture)
RsquareAdj(vector_moisture_cca, 100)
anova(vector_moisture_cca, permutations = 999)
anova(vector_moisture_cca, by='margin', permutations = 999)

par(mfrow=(c(1,2)))
plot(vector_moisture_cca, type='n', scaling=1)
orditorp(vector_moisture_cca, display='sp', cex=0.5, scaling=1, col='blue')
text(vector_moisture_cca, display='bp', col='red')


```

vector manure

```{r}
vector_manure <- as.numeric(as.character(dune.env$Manure))
vector_manure_cca <- cca(dune~vector_manure)
RsquareAdj(vector_manure_cca, 100)
anova(vector_manure_cca, permutations = 999)
anova(vector_manure_cca, by='margin', permutations = 999)

plot(vector_manure_cca, type='n', scaling=1)
orditorp(vector_manure_cca, display='sp', cex=0.5, scaling=1, col='blue')
text(vector_manure_cca, display='bp', col='red')
```
new dune cca
```{r}
cca_dune_new <- cca(dune ~ A1 + Use + Management + vector_moisture + vector_manure, data=dune.env)
RsquareAdj(cca_dune_new, 100)
anova(cca_dune_new, permutations = 999)
anova(cca_dune_new, by='margin', permutations = 999)

plot(cca_dune_new, type='n', scaling=1)
orditorp(cca_dune_new, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune_new, display='cn', col='red')
```
# 3) Do your two analyses agree with one another or complement one another or do these two analyses seem to be suggesting different take home messages? Which analysis do you find to be more useful?

```{r}
par(mfrow=(c(1,3)))
plot(vector_moisture_cca, type='n', scaling=1)
orditorp(vector_moisture_cca, display='sp', cex=0.5, scaling=1, col='blue')
text(vector_moisture_cca, display='bp', col='red')

# compared with the nmds moisture graph
plot(dune_mds, type='n')
text(dune_mds, 'sp', cex=.5)
# generate vector of colors 
color_vect <- rev(terrain.colors(6))[-1]
points(dune_mds, 'sites', pch=19, 
       col=color_vect[dune.env$Moisture])
legend('topright', paste("Moisture =", 1:5, sep=''), 
       col=color_vect, pch=19)
# also compared with overall cca analysis 
plot(cca_dune_new, type='n', scaling=1)
orditorp(cca_dune_new, display='sp', cex=0.5, scaling=1, col='blue')
text(cca_dune_new, display='cn', col='red')
```

As seen above the CCA analysis of vector moisture and the NMDS analysis do seem to reperesent the same conclusions regarding moisture. This can be seen be the high positive correlation of Eleopaulu, Ranuflam, and Juncarti in the graphic of the CCA analysis, as well as their correspondence with moisture level five, a higher moisture level, shown by the NMDS analysis. This is further shown by the begative correlation of Tripfrat and Achmill in the CCA graph, as well as their correspondence with moisture level 1 and 2 in the nmds graph. Lastly, the high residual of Airaprae and Empenigr in the CCA graph coincides with their lack of correspondence with any moisture levels in the NMDS graph. This shows that there not much variance of these species is explained by moisture, and they are most likely affected more by other variables in the model. 

The two analyses do agree in the sense that moisture is definitely one of the most important factorial variables in this study. However, we found the CCA analysis to be slightly more useful in determining a plot that would include other variables that are clearly of importance such as manur and management. The CCA model also shows relationships among explanatory variables with one another and with the species represented along an axis, while the NMDS model is more limited to showing the observation of moisture in relation to each species. In regards to moisture, both analyses share a similar take-home message.

# further CCA analysis 

```{r}
RsquareAdj(cca_dune_new, 100)
RsquareAdj(cca_dune_A1, 100)
RsquareAdj(cca_dune_Management, 100)
RsquareAdj(cca_dune_Use, 100)
RsquareAdj(cca_dune_moisture, 100)
RsquareAdj(cca_dune_Manure, 100)
RsquareAdj(vector_manure_cca, 100)
RsquareAdj(vector_moisture_cca, 100)
```
should try removing use from model, only 1.6 rsquared adj
chnaging to vectors, although better to visualize does decrease the r square, but increases adj r squre,
so overall probably a better model even though individuallly the vectors have lower r squres than their counterparts
they help to decrease the overfitting of the model due to an excess of variables
```{r}
without_use_cca <- update(cca_dune_new, . ~ . - Use)
par(mfrow=c(1,1))
RsquareAdj(without_use_cca, 100)
anova(without_use_cca, permutations = 999)
anova(without_use_cca, by='margin', permutations = 999)

plot(without_use_cca, type='n', scaling=1)
orditorp(without_use_cca, display='sp', cex=0.5, scaling=1, col='blue')
text(without_use_cca, display='cn', col='red')
```
does increase the adj r but only by 1%, overall a very similar looking model 


manure plot like moisture nmds
```{r}
plot(dune_mds, type='n')
text(dune_mds, 'sp', cex=.5)
# generate vector of colors 
color_vect = rev(heat.colors(6))[-1]
points(dune_mds, 'sites', pch=19, 
       col=color_vect[dune.env$Manure])
legend('topright', paste("Manure =", 1:5, sep=''), 
       col=color_vect, pch=19)
# dune.env$Manure
```
this further shows the correlation between the cca and nmds ordination results. 



